Spike spectra for recurrences

### 2022-11-04 Manuscript/inter_spike_manuscript_mdpi.tex
rebuttal letter

Recurrence plots (RPs) provide a vivid representation of complex dynamics $\vec{x}_i$ stemming from potentially high dimensional systems \cite{marwan2007}
\begin{equation}\label{eq_rp_definition}
R_{i,j}(\varepsilon) = \Theta\left(\varepsilon - D_{i,j}\right) 
= \Theta\left(\varepsilon - \| \vec{x}_i - \vec{x}_j\|\right), \qquad \vec{x} \in \mathbb{R}^d, \quad i,j \in [1,\ldots, N],
\end{equation}
with $\mathbf{R}$ the recurrence matrix, satisfactory$\vec{x}_i$ the state vector at time 
$t = \Delta t \cdot i$ ($\Delta t$ the sampling time), $N$ the number of
sampling points (or length of data series), and $d$ the dimension of the system.
The crucial free parameter $\varepsilon$ is the recurrence threshold, determining what is a recurrences
and, thus, the visible structures in the RP. It can be chosen such that the recurrence rate 
$RR(\varepsilon)=N^{-2}\sum_{i,j}^N R_{i,j}(\varepsilon)$ exhibits a certain value \cite{kraemer2018}.
The simple idea to track recurring states of the $d$-dimensional trajectory $\vec{x}_i$ of the system under study not only allows for a beneficial visualization of the dynamics, 
but also for its 
quantification, using certain structures in the RP, such as diagonal or vertical lines \cite{marwan2007}.

@@

\citet{Zbilut2008} pointed out that $\tau$-RR could be interpreted as analogous to the auto-correlation function $C(\tau)$ for a scalar time series.
However, as there is no equivalent of the Wiener-Khinchin theorem its Fourier Transform (FT) only provides a ``generalized'' kind of power spectral density.
There are clear advantages to such a recurrence-derived power spectral density, the (FT) $\tau$-RR (Fig.~\ref{fig_tau_rr_spectrum_example}D,E):
There are no assumptions for stationarity or sampling, when constructing a RP, 
and since an RP can represent high dimensional dynamics the approch directly applies to higher dimensional systems for which the auto-correlation is a matrix and no single power spectrum captures the recurrences.\\

\textbf{Please check the above paragraph for correctness, I did not understand the following and would drop it (or explain the point better):}
\textit{This is reasonable, since the average distances for a given lag $\tau$ 
\begin{equation}
\overline{D}(\tau) = \frac{1}{N-\tau}\sum_{i=1}^{N-\tau} D_{i, i+\tau}
\end{equation}
can be directly read from the distance matrix $\mathbf{D}$ and is also preserved in its thresholded version $\tau$-RR.}

@@

However, the exact meaning of this generalized power spectral density is unclear, and it is typically hard to interpret. $\tau$-RR spikes already directly encode the periodicity of the underlying signal, in contrast it is unclear what, if any, interpretation the "power" contained in a particular frequency mode of the $\tau$-RR should have. \textbf{In particular it does not imply anything about the power associated to certain frequencies in the underlying timeseries.} Whenever $\tau$-RR is a spike-train-like signal, which it is in most cases (see Fig.~\ref{fig_tau_rr_spectrum_example}) especially for 
map-data (low-resolution data), a FT of such a signal leads to a spike-train-like image in the frequency domain (e.g., \cite{Schild1982,Cordoba1989}, see 
Fig.~\ref{fig_tau_rr_spectrum_example}E). Thus, it is not intuitive how to extract meaningful information about the systems' state space trajectory.

@@

In this article we propose a new way of decomposing a spike-train-like signal into periodic components as 
an alternative to the RP-based method suggested in \cite{Zbilut2008} or Fourier-based spike-train power 
spectra \cite{Dummer2014}. We instead investigate decomposing the $\tau$-RR in an overcomplete basis of spike trains using a sparsity regularization. Section \ref{sec_tau_rr_method} introduces the approach in detail. Our novel \textit{inter-spike spectrum} does not show resonance 
behavior of the signal's inherent inter-spike frequencies (Fig.~\ref{fig_tau_rr_dirac_comb}C, F). Note that this approach can be used to decompose arbitrary signals, and is not specific to $\tau$-RR. The more spiky the signal is, the more useful our new decomposition can potentially be. In Section \ref{sec_tau_rr_application} we demonstrate the usage of inter-spike spectrum when transforming the $\tau$-RR of a system under study. In this case, the inter-spike spectrum 
can unravel characteristic time scales of high dimensional systems, which is not possible when using a FT. Finally, in Section \ref{sec_tau_rr_conclusion} our results are summarized.

@@

To obtain the Inter-Spike Spectrum , the signal, which we would like to analyze, in our case the $\tau$-RR, is decomposed into a set of appropriate basis functions. 
The general idea common to many methods is that the sum of these weighted functions can approximate a finite signal to a sufficient degree. The weights (in some contexts also called modes or loadings) corresponding to the individual basis functions must be determined.
A number of decomposition techniques based on different sets of basis functions exist, e.g., 
trigonometric functions (Fourier- and wavelet analysis \cite{Bracewell1986}), 
eigenvectors of the corresponding covariance matrix (Principal Component Analysis \cite{Hotelling1933} and related techniques) 
or Intrinsic Mode Functions (Empirical Mode Decomposition and Hilbert spectrum \cite{Norden1998}).
These methods typically share the property that the basis functions form a complete basis,
the set of basis functions is linearly independent, and thus the weights are uniquely determined.

>>>>

Recurrence plots (RPs) provide a vivid representation of complex dynamics $\vec{x}_i$ stemming from potentially high dimensional systems \cite{marwan2007}
\begin{equation}\label{eq_rp_definition}
R_{i,j}(\varepsilon) = \Theta\left(\varepsilon - D_{i,j}\right) 
= \Theta\left(\varepsilon - \| \vec{x}_i - \vec{x}_j\|\right), \qquad \vec{x} \in \mathbb{R}^d, \quad i,j \in [1,\ldots, N],
\end{equation}
with $\mathbf{R}$ the recurrence matrix, satisfactory $\vec{x}_i$ the state vector at time 
$t = \Delta t \cdot i$ ($\Delta t$ the sampling time), $N$ the number of
sampling points (or length of data series), and $d$ the dimension of the system.
The crucial free parameter $\varepsilon$ is the recurrence threshold, determining what is a recurrences
and, thus, the visible structures in the RP. It can be chosen such that the recurrence rate 
$RR(\varepsilon)=N^{-2}\sum_{i,j}^N R_{i,j}(\varepsilon)$ exhibits a certain value \cite{kraemer2018}.
The simple idea to track recurring states of the $d$-dimensional trajectory $\vec{x}_i$ of the system under study not only allows for a beneficial visualization of the dynamics, 
but also for its 
quantification, using certain structures in the RP, such as diagonal or vertical lines \cite{marwan2007}.

@@

\citet{Zbilut2008} pointed out that $\tau$-RR could be interpreted as analogous to the auto-correlation function $C(\tau)$ for a scalar time series.
However, as there is no equivalent of the Wiener-Khinchin theorem its Fourier Transform (FT) only provides a ``generalized'' kind of power spectral density.
There are clear advantages to such a recurrence-derived power spectral density, the (FT) $\tau$-RR (Fig.~\ref{fig_tau_rr_spectrum_example}D,E):
There are no assumptions for stationarity or sampling, when constructing a RP, 
and since an RP can represent high dimensional dynamics the approch directly applies to higher dimensional systems for which the auto-correlation is a \todo{hm, I'm not so happy with the "matrix" here} matrix and no single power spectrum captures the recurrences.

\todo[inline]{I agree that the following is not necessary and could be removed}
\textbf{Please check the above paragraph for correctness, I did not understand the following and would drop it (or explain the point better):}
\textit{This is reasonable, since the average distances for a given lag $\tau$ 
\begin{equation}
\overline{D}(\tau) = \frac{1}{N-\tau}\sum_{i=1}^{N-\tau} D_{i, i+\tau}
\end{equation}
can be directly read from the distance matrix $\mathbf{D}$ and is also preserved in its thresholded version $\tau$-RR.}

@@

However, the exact meaning of this generalized power spectral density is unclear, and it is typically hard to interpret. $\tau$-RR spikes already directly encode the periodicity of the underlying signal, in contrast it is unclear what, if any, interpretation the ``power'' contained in a particular frequency mode of the $\tau$-RR should have. \textbf{In particular it does not imply anything about the power associated to certain frequencies in the underlying timeseries.} Whenever $\tau$-RR is a spike-train-like signal, which it is in most cases (see Fig.~\ref{fig_tau_rr_spectrum_example}) especially for 
map-data (low-resolution data), a FT of such a signal leads to a spike-train-like image in the frequency domain (e.g., \cite{Schild1982,Cordoba1989}, see 
Fig.~\ref{fig_tau_rr_spectrum_example}E). Thus, it is not intuitive how to extract meaningful information about the systems' state space trajectory.

@@

In this article we propose a new way of decomposing a spike-train-like signal into periodic components as 
an alternative to the RP-based method suggested in \cite{Zbilut2008} or Fourier-based spike-train power 
spectra \cite{Dummer2014}. We instead investigate decomposing the $\tau$-RR in an overcomplete basis of spike trains using a sparsity regularization. Section \ref{sec_tau_rr_method} introduces the approach in detail. Our novel \textit{inter-spike spectrum} does not show resonance 
behavior of the signal's inherent inter-spike frequencies (Fig.~\ref{fig_tau_rr_dirac_comb}C, F). This approach can be used to decompose arbitrary signals, and is not specific to $\tau$-RR. The more spiky the signal is, the more useful our new decomposition can potentially be. In Section \ref{sec_tau_rr_application} we demonstrate the usage of inter-spike spectrum when transforming the $\tau$-RR of a system under study. In this case, the inter-spike spectrum 
can unravel characteristic time scales of high dimensional systems, which is not possible when using a FT. Finally, in Section \ref{sec_tau_rr_conclusion} our results are summarized.

@@

To obtain the Inter-Spike Spectrum , the signal, which we would like to analyze, in our case the $\tau$-RR, is decomposed into a set of appropriate basis functions. 
The general idea common to many methods is that the sum of these weighted functions can approximate a finite signal to a sufficient degree. The weights (in some contexts also called modes or loadings) corresponding to the individual basis functions must be determined.
A number of decomposition techniques based on different sets of basis functions exist, e.g., 
trigonometric functions (Fourier- and wavelet analysis \cite{Bracewell1986}), 
eigenvectors of the corresponding covariance matrix (principal component analysis \cite{Hotelling1933} and related techniques) 
or intrinsic mode functions (empirical mode decomposition and Hilbert spectrum \cite{Norden1998}).
These methods typically share the property that the basis functions form a complete basis,
the set of basis functions is linearly independent, and thus the weights are uniquely determined.
